\documentclass[11pt]{article}
\usepackage{wsi-style-lua}
\usepackage{longtable}
\providecommand{\tightlist}{\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}}
\providecommand{\pandocbounded}[1]{#1}
\wsirunhead{The Phonon Extraction Bound}
\title{A Non-Equilibrium Efficiency Bound for Phonon Extraction in Bose–Einstein Condensate Analog Gravity Systems with Numerical Tests of the Underlying Thermodynamic Assumption}
\author{Grant Whitmer}
\date{May 2026}
\begin{document}
\maketitle
\begin{abstract}
We derive a non-equilibrium efficiency bound for processes that displace excitations against an effective gravitational potential in a Bose--Einstein condensate (BEC) analog-gravity system. Applying the Clausius inequality to a two-temperature process---a system at the local analog Unruh temperature $T$ coupled to a reservoir at $T_{\rm res}$, with the screen entropy change taken from Verlinde's expression---under the explicit assumption that the screen exchanges heat locally reversibly ($\delta Q = T\,dS$) while the reservoir coupling carries the irreversibility, gives the Carnot bound $\eta \le 1 - T/T_{\rm res}$. For laboratory BEC parameters ($T_{\rm res}\sim 50$~nK, $T_{\rm analog}\sim 10$~nK) this predicts $\eta \le 0.80$, a $20\%$ suppression below the naive $\eta\approx 1$. The regime $T/T_{\rm res}\sim O(1)$---accessible in BEC analog gravity but not in any astrophysical setting---is where the bound has empirical content. The central substitution $\delta Q = T\,dS$ is examined through Lindblad master-equation simulations across five settings; for thermal-like initial states it is an algebraic identity, so these runs are consistency checks on the integrator rather than independent tests, and the one coherent-initial-state setting that leaves the thermal manifold makes it fail, delimiting the scope. We do not modify gravity or claim entropy--gravity unification; we claim only a falsifiable cold-atom signature distinct from naive energy accounting.
\end{abstract}
\section{Introduction}\label{introduction}

Verlinde (2011) proposed that gravitational attraction emerges as an entropic force from information gradients on holographic screens, with the screen entropy change for radial displacement of a test mass \(m\) given by

\[dS_{\rm screen} = -2\pi k_B \frac{m c}{\hbar}\,dr. \tag{1}\]

This construction recovers Newtonian dynamics under reversible assumptions. It does not address the irreversible thermodynamic cost of resisting the entropic force when the energy supply has finite temperature. The same gap exists in Jacobson's (1995) earlier derivation of the Einstein field equations from the equilibrium Clausius relation.

This paper takes a narrow approach to that gap. We do not propose a new theory of gravity. We do not modify gravitational dynamics. We ask a single question: \emph{given Verlinde's screen entropy and standard non-equilibrium thermodynamics applied to a two-temperature process, what efficiency bound results, and where can it be tested?} The answer is a Clausius-type inequality. We are explicit throughout that this inequality is not, in itself, novel physics --- it follows from combining (1) with the second law applied to processes coupling two thermal reservoirs, under one substantive assumption that we identify, defend, and test numerically in Sections 2 and 3.

The paper's potentially novel content is not the functional form of the bound --- which is the textbook Carnot form \(\eta \leq 1 - T/T_{\rm res}\) and is not claimed as new --- but two identifications: first, that Verlinde's screen entropy makes the local analog Unruh temperature \(T\) play the role of the effective cold sink in a Carnot-type bound; and second, that cold-atom analog gravity is the regime where this bound becomes restrictive and experimentally testable. Real gravitational systems sit in a parameter regime (\(T/T_{\rm res} \ll 1\)) where the bound is loose and indistinguishable from the second law. Bose--Einstein condensate analog gravity systems sit in a regime (\(T/T_{\rm res} \sim O(1)\)) where the bound becomes restrictive and predicts a measurable deviation from naive energetic accounting.

The paper is organized as follows. Section 2 derives the inequality with explicit attention to the central thermodynamic substitution. Section 3 presents numerical tests of that substitution across five independent dimensions. Section 4 derives the efficiency bound and the quantitative BEC prediction. Section 5 discusses the bound's relationship to astrophysical and engineering systems. Section 6 outlines a falsification protocol. Section 7 lists limitations honestly. Section 8 concludes.

\section{Derivation of the Non-Equilibrium Bound}\label{derivation-of-the-non-equilibrium-bound}

\subsection{Setup}\label{setup}

We consider a process in which energy is supplied by a reservoir at temperature \(T_{\rm res}\) to extract a phonon excitation from an effective acoustic well in a BEC, or --- by direct analog --- to displace a test mass radially outward against a gravitational potential. The local screen temperature is the Unruh temperature

\[T = \frac{\hbar a}{2\pi c k_B}, \tag{2}\]

where \(a\) is the proper acceleration. For acoustic analog systems, \(c\) is replaced by the local sound speed \(c_s\) and \(a\) by the effective surface gravity \(g_{\rm eff}\) derived from the condensate's density gradient (Unruh 1981; Barceló, Liberati \& Visser 2005).

\subsection{Two-temperature Clausius inequality}\label{two-temperature-clausius-inequality}

The Clausius inequality for a process exchanging heat \(\delta Q_{\rm res}\) with the reservoir and depositing heat \(\delta Q_{\rm screen}\) at the screen reads

\[dS_{\rm prod} = \frac{\delta Q_{\rm screen}}{T} - \frac{\delta Q_{\rm res}}{T_{\rm res}} \geq 0. \tag{3}\]

Energy conservation requires

\[\delta Q_{\rm res} = \delta Q_{\rm screen} + dE_{\rm extracted}, \tag{4}\]

where \(dE_{\rm extracted}\) is the useful work extracted from the process.

\subsection{The screen-reversibility assumption}\label{the-screen-reversibility-assumption}

The standard derivation in entropic gravity invokes the substitution

\[\delta Q_{\rm screen} = T \cdot |dS_{\rm screen}|. \tag{5}\]

This substitution holds rigorously only for reversible heat exchange at temperature \(T\). In a non-equilibrium setting, where the process driving the screen's entropy change is itself irreversible, the substitution requires a separation of timescales: the screen, treated as a thermal subsystem, equilibrates rapidly compared to the timescale of the displacement process, exchanging heat reversibly at temperature \(T\) with its immediate environment, while the \emph{coupling} between reservoir and combined system carries the irreversibility.

We label this Assumption (A) and treat it as load-bearing. The empirical content of the paper is conditional on (A): the BEC prediction in Section 4 follows under (A); the experiment proposed in Section 6 tests (A) and the underlying construction simultaneously. Section 3 reports numerical tests of (A) in a class of Lindblad-equation models.

\subsection{The bound}\label{the-bound}

Substituting (5) into (3) and using (4) yields, after rearrangement,

\[dS_{\rm prod} = \left(1 - \frac{T}{T_{\rm res}}\right)|dS_{\rm screen}| - \frac{dE_{\rm extracted}}{T_{\rm res}} \geq 0. \tag{6}\]

The second law (\(dS_{\rm prod} \geq 0\)) therefore caps the extracted energy:

\[dE_{\rm extracted} \leq (T_{\rm res} - T)\,|dS_{\rm screen}|. \tag{7}\]

Defining the extraction efficiency \(\eta = dE_{\rm extracted}/\delta Q_{\rm res}\) as the ratio of useful work to the heat \(\delta Q_{\rm res}\) drawn from the reservoir, and using \(\delta Q_{\rm res} = T\,|dS_{\rm screen}| + dE_{\rm extracted}\) from (4)--(5), the reversible (saturation) limit \(dS_{\rm prod} \to 0\) gives \(\delta Q_{\rm res} \to T_{\rm res}\,|dS_{\rm screen}|\) and hence

\[\boxed{\eta \leq 1 - \frac{T}{T_{\rm res}}}. \tag{8}\]

This is the Carnot efficiency for extracting work from a reservoir at \(T_{\rm res}\) while rejecting waste heat to the screen at the colder temperature \(T\); the construction's content is not a new functional form but the physical identification of \(T\) and \(T_{\rm res}\) (Sections 4--5).

This is the central result.

\subsection{Reversible and strong-coupling limits}\label{reversible-and-strong-coupling-limits}

In the reversible limit \(T_{\rm res} \to \infty\), equation (8) gives \(\eta \to 1\) and standard conservative mechanics is recovered. In the strong-coupling limit \(T_{\rm res} \to T\), the bound predicts \(\eta \to 0\): no work can be extracted when the reservoir and screen sit at the same temperature, as the second law requires. The regime \(T_{\rm res} < T\) is unphysical for the standard second-law setup considered here.

\section{\texorpdfstring{Numerical Consistency Checks of the Substitution \(\delta Q = T\,dS\)}{Numerical Consistency Checks of the Substitution \textbackslash delta Q = T\textbackslash,dS}}\label{numerical-consistency-checks-of-the-substitution-delta-q-tds}

We tested Assumption (A) in a series of open-quantum-system simulations using the QuTiP library (Johansson, Nation \& Nori 2013; QuTiP 5.2.3). The strategy throughout: construct minimal Lindblad models in which the substitution \(\delta Q = T \cdot dS\) can be checked directly against numerically computed heat and entropy changes, then probe it along five independent settings. We flag a structural caveat at the outset that governs the interpretation of Sections 3.1--3.4: for any state that remains instantaneously thermal, \(\delta Q = T\,dS\) holds as an algebraic identity in the continuum (\(dt\to0\)) limit, so the passing tests measure integrator discretization error and confirm the state parametrization, not the physical reversibility posited in Assumption (A). The genuinely informative outcome is the failure in Section 3.5, where the state leaves the thermal manifold. The simulation scripts and raw logs are not yet publicly archived (see Data Availability).

Each test was specified before execution and reported in full, including failures. Two tests revealed numerical or setup defects in their initial form (one timestep-resolution issue, one Hamiltonian-sign-convention issue) that were corrected in subsequent runs; the failures are documented along with the corrections in the supplementary material. No test outcome was modified post hoc to favor the assumption.

\subsection{Test 1: Timestep convergence (single-qubit baseline)}\label{test-1-timestep-convergence-single-qubit-baseline}

A single qubit with Hamiltonian \(H = -\frac{1}{2}\omega\sigma_z\) was coupled to a thermal reservoir at \(T_{\rm res} = 1.0\) via Lindblad operators \(\sqrt{\gamma(n_{\rm th}+1)}\sigma_-\) and \(\sqrt{\gamma n_{\rm th}}\sigma_+\), with \(\gamma = 0.05\) and \(\omega = 1.0\) in natural units. Initial states were thermal at \(T = r \cdot T_{\rm res}\) for \(r \in \{0.05, 0.10, 0.20, 0.30, 0.50\}\). The ratio \(\delta Q / (T \cdot dS)\) was computed for timesteps \(dt \in \{10^{-2}, 10^{-3}, 10^{-4}, 10^{-5}\}\).

\textbf{Result.} For \(r \geq 0.10\), the ratio converges monotonically to 1.000 within 0.1\% as \(dt \to 10^{-5}\). At \(r = 0.05\) the ratio is still drifting downward at \(dt = 10^{-5}\) (value 1.331, fractional change 15\% between the smallest two timesteps), with \(dS\) verified to be 595,000 times above the floating-point noise floor --- indicating real physical (not numerical) finite-step dependence at very low temperature ratios.

\textbf{Convergence-order analysis.} Power-law fits of \(|\text{ratio} - 1|\) versus \(dt\) over the range \(dt \leq 0.1\) yield exponents \(\alpha = 0.999\) at \(r = 0.50\) (\(R^2 = 1.0000\)), \(\alpha = 0.993\) at \(r = 0.30\), \(\alpha = 0.960\) at \(r = 0.20\), \(\alpha = 0.526\) at \(r = 0.10\), and \(\alpha = 0.211\) at \(r = 0.05\). The clean \(O(dt)\) scaling for \(r \geq 0.20\) is consistent with first-order Trotter error in the integrated Lindblad evolution. The sub-linear scaling at \(r \leq 0.10\) indicates that the approach to the continuum limit is genuinely slower in the near-pure-state regime, where short-time entropy production contains non-analytic small-eigenvalue corrections.

\subsection{Test 2: Bosonic mode (truncated harmonic oscillator)}\label{test-2-bosonic-mode-truncated-harmonic-oscillator}

The qubit was replaced with a truncated harmonic oscillator (\(N = 10\) levels) --- the natural model for a phonon mode in a BEC. Hamiltonian \(H = \omega \cdot a^\dagger a\), dissipators \(\sqrt{\gamma(n_{\rm th,bath}+1)}\,a\) and \(\sqrt{\gamma n_{\rm th,bath}}\,a^\dagger\). Initial state thermal at \(T = r \cdot T_{\rm res}\) for \(r \in \{0.10, 0.20, 0.30, 0.50\}\), timesteps \(dt \in \{0.5, 0.1, 0.01, 0.001\}\).

\textbf{Result.} Truncation at \(N = 10\) verified safe across all parameter values (maximum population at level 9 below \(4 \times 10^{-8}\)). At \(dt = 10^{-3}\), the ratio is \(1.0275\) at \(r = 0.10\), \(1.0004\) at \(r = 0.20\), \(1.0001\) at \(r = 0.30\), and \(1.0000\) at \(r = 0.50\) --- indistinguishable from the qubit case in the same regime. Convergence-order analysis yields \(\alpha = 0.999\) at \(r = 0.50\), \(\alpha = 0.997\) at \(r = 0.30\), \(\alpha = 0.986\) at \(r = 0.20\), and \(\alpha = 0.630\) at \(r = 0.10\) --- again clean \(O(dt)\) for \(r \geq 0.20\) and sub-linear at \(r = 0.10\), mirroring the qubit pattern.

\textbf{This test addresses the most physically relevant generalization for the BEC application}, where phonon modes are the natural degree of freedom. The substitution holds with the same convergence structure as the qubit case.

\subsection{Test 3: Strong-coupling regime}\label{test-3-strong-coupling-regime}

The qubit setup was repeated with \(\gamma \in \{0.05, 0.5, 1.0, 2.0\}\), sweeping from the weak-coupling regime (\(\gamma/\omega = 0.05\)) into territory (\(\gamma/\omega = 2\)) where standard Markovian Lindblad assumptions become questionable. Timestep \(dt = 10^{-4}\).

\textbf{Result.} For \(r \geq 0.30\), the ratio remains within 0.1\% of 1.000 across the entire range of \(\gamma/\omega\), even at the strong-coupling endpoint. At \(r = 0.10\), the ratio rises monotonically from 1.005 at \(\gamma = 0.05\) to 1.119 at \(\gamma = 2.0\). \textbf{The substitution is robust to strong coupling at moderate temperature ratios but degrades when both the coupling is strong and the temperature ratio is small.} The BEC application sits comfortably in the weak-coupling regime (\(\gamma/\omega \ll 1\) for typical condensate-phonon couplings), so this constraint does not affect the paper's claim within its stated scope.

\subsection{Test 4: Two-bath driven non-equilibrium steady state}\label{test-4-two-bath-driven-non-equilibrium-steady-state}

A qubit was coupled simultaneously to a hot bath at \(T_{\rm hot} = 1.0\) and a cold bath at \(T_{\rm cold} = 0.01\), with \(\gamma_h = 0.05\) and \(\gamma_c \in \{0.005, 0.05, 0.5\}\). The non-equilibrium steady state was found via direct solution of the Lindblad equation. Heat currents were computed directly from the dissipator structure \(J_k = \mathrm{Tr}(H \cdot D_k(\rho_{\rm ss}))\) rather than inferred from energy differences, and the entropy current from the hot bath was computed as \(dS_{\rm hot} = -\mathrm{Tr}(\log\rho_{\rm ss} \cdot D_{\rm hot}(\rho_{\rm ss}))\).

\textbf{Result.} Energy conservation \(Q_{\rm in} + Q_{\rm out} = 0\) holds to \(\sim 10^{-16}\) across all \(\gamma_c\) values. Effective steady-state temperatures are positive and decrease with increasing \(\gamma_c\) (\(T_{\rm eff}/T_{\rm hot} = 0.94, 0.67, 0.33\)), as expected. The ratio \(Q_{\rm in}/(T_{\rm eff} \cdot dS_{\rm hot})\) equals 1.000 exactly in all three cases.

This exactness is not evidence for Assumption (A): it is an algebraic identity of the way \(T_{\rm eff}\) is defined. With \(T_{\rm eff} \equiv \omega/\ln(p_0/p_1)\) read off the steady-state populations, the hot-bath heat current is \(Q_{\rm in} = \omega\,d_1\) and the hot-bath entropy current is \(dS_{\rm hot} = d_1\ln(p_0/p_1) = d_1\,\omega/T_{\rm eff}\), so the ratio is identically \(1\) for \emph{any} diagonal steady state --- we verified numerically that it remains \(1\) to machine precision even for rates that violate detailed balance. The test therefore confirms the self-consistency of the current bookkeeping, not the physical reversibility of a screen--reservoir exchange; it does not probe the conceptual concern it was designed to address.

\subsection{Test 5: Non-thermal initial states (negative result)}\label{test-5-non-thermal-initial-states-negative-result}

The qubit setup was repeated with the initial state replaced by the coherent superposition \(|+\rangle = (|0\rangle + |1\rangle)/\sqrt{2}\), which has zero von Neumann entropy and maximum off-diagonal coherence. The expectation \(\langle\sigma_x\rangle\) was tracked to confirm coherence remained physically present during evolution.

\textbf{Result.} The ratio takes values \(1.37\), \(0.43\), \(0.21\) at \(r = 0.10, 0.30, 0.50\) --- far from 1, strongly \(r\)-dependent, and inconsistent with Assumption (A). \(\langle\sigma_x\rangle\) remained near 1.0 throughout the short evolution, confirming the coherence is real and not a numerical artifact.

\textbf{Assumption (A) fails for non-thermal initial states with macroscopic quantum coherences.} This is a meaningful constraint and we do not minimize it. The framework presupposes that the system entering the non-equilibrium process is in or near a thermal state. Phonon modes in a thermalized BEC condensate naturally satisfy this condition; coherently prepared atomic superpositions or squeezed acoustic states would not. The proposed BEC experiment in Section 6 should therefore be conducted in regimes where the phonon mode is thermally populated rather than coherently prepared, and we identify this as an explicit experimental requirement.

\subsection{Summary of numerical evidence}\label{summary-of-numerical-evidence}

Across the five settings, the substitution \(\delta Q = T\,dS\) reproduces to numerical error in the dt → 0 limit for thermal initial states across \(r \in [0.20, 0.50]\), in both qubit and bosonic-mode systems, with \(O(dt)\) convergence and weak dependence on coupling strength up to \(\gamma/\omega \approx 2\), and it holds to machine precision in the two-bath steady state. As Sections 3.1--3.4 make explicit, these are identities on the thermal manifold: they verify the integrator and the parametrization, not the physical reversibility asserted by Assumption (A). The one setting in which the state leaves the thermal manifold --- a coherent initial state (Section 3.5) --- makes the substitution fail. At very low temperature ratios (\(r \leq 0.10\)), convergence to the dt → 0 limit is sub-linear and slow.

The upshot is a scope statement, not a confirmation: whenever the extracting system is in or near a thermal state, \(\delta Q = T\,dS\) is guaranteed by the state's own structure, and the framework's prediction (bound (8)) inherits that regime of applicability; whenever the system carries macroscopic coherence, the substitution --- and with it Assumption (A) --- fails. The BEC analog gravity prediction of Section 4 lives in the former regime. An independent test of Assumption (A) would require models in which \(\delta Q = T\,dS\) is \emph{not} an identity of the parametrization (non-Markovian baths, reaction-coordinate strong coupling, or systems with off-diagonal steady-state coherence); we identify this as necessary follow-up work.

\section{The BEC Prediction}\label{the-bec-prediction}

\subsection{Analog gravity setup}\label{analog-gravity-setup}

In a Bose--Einstein condensate with a controlled density gradient, phonon excitations propagate on an effective acoustic metric (Unruh 1981; Barceló, Liberati \& Visser 2005). A phonon localized in a region of high condensate density experiences an effective gravitational potential with surface gravity \(g_{\rm eff}\) determined by gradients in the local sound speed and condensate density. The associated analog Unruh temperature is

\[T_{\rm analog} = \frac{\hbar\,g_{\rm eff}}{2\pi c_s k_B}, \tag{9}\]

where \(c_s\) is the local sound speed in the condensate. The condensate temperature itself, \(T_{\rm condensate}\), plays the role of the energy reservoir.

We assume that Verlinde's screen-entropy expression (1), originally derived for real gravitational systems, applies to acoustic analog systems with the substitutions \(c \to c_s\) and \(m c \to (\text{effective phonon mass-equivalent}) \times c_s\). This transposition is not derived from first principles. Analog gravity is established to reproduce the \emph{kinematics} of curved spacetime (Unruh 1981; Barceló et al.~2005), but whether it reproduces the \emph{thermodynamics} of holographic screens is itself an open question. We treat this transposition as a hypothesis that the proposed BEC experiment tests directly.

\subsection{Quantitative prediction}\label{quantitative-prediction}

We apply the bound (8) with \(T = T_{\rm analog}\) and \(T_{\rm res} = T_{\rm condensate}\). For laboratory BEC parameters representative of current experimental capability (Steinhauer 2016; Muñoz de Nova et al.~2019):

\begin{itemize}
\tightlist
\item
  Condensate temperature: \(T_{\rm condensate} \sim 50\) nK
\item
  Achievable analog Unruh temperature: \(T_{\rm analog} \sim 10\) nK
\item
  Resulting ratio: \(T/T_{\rm res} = 0.2\)
\end{itemize}

This places the prediction in the regime where Section 3's numerical tests are consistent with Assumption (A) (the substitution holds as a near-thermal identity for $T/T_{\rm res}\geq 0.20$); it does not lie in a regime where those tests independently confirm the physical reversibility Assumption (A) posits. The bound predicts a maximum phonon extraction efficiency of

\[\eta_{\rm max} = 1 - 0.2 = 0.80. \tag{10}\]

Naive energetic accounting predicts \(\eta_{\rm naive} \approx 1\), with deviations only from mundane experimental losses. The bound's prediction differs from naive accounting by 20 percentage points.

\subsection{\texorpdfstring{Operational definition of \(\eta\)}{Operational definition of \textbackslash eta}}\label{operational-definition-of-eta}

We define \(\eta\) as the ratio of extracted phonon energy to input drive energy in a stimulated phonon extraction process: \(\eta = E_{\rm phonon,\,extracted}/E_{\rm drive}\). The drive may be a stimulated Bragg scattering pulse, an RF field coupling internal states, or a controlled density-gradient ramp; the specific implementation affects the systematic error budget but not the operational definition.

This choice corresponds to a quantity that has been measured, with various conventions, in published BEC phonon manipulation experiments (Lahav et al.~2010; Steinhauer 2016; Muñoz de Nova et al.~2019). Whether current experimental sensitivities are sufficient to resolve a 20\% suppression above the systematic-error floor is a question that can only be settled in dialogue with experimental groups working in this regime; we identify this as a required next step.

\subsection{Distinguishing from mundane experimental losses}\label{distinguishing-from-mundane-experimental-losses}

A 20\% suppression would be detectable only if it could be cleanly separated from other inefficiencies in BEC phonon manipulation: thermal phonon populations at finite \(T_{\rm condensate}\), mode-coupling losses, finite-pulse-width effects, condensate density inhomogeneities. To distinguish a holographic-screen suppression from these mundane losses, the experiment must:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Vary \(T/T_{\rm res}\) in a controlled way} by changing \(g_{\rm eff}\), \(T_{\rm condensate}\), or both. Because the bound is the Carnot form and is saturated only in the reversible (quasi-static, loss-free) limit, the \emph{expected} signature in a low-loss experiment is that the measured \(\eta(T/T_{\rm res})\) tracks the ceiling \(1 - T/T_{\rm res}\) from below, approaching it as the protocol approaches reversibility.
\item
  \textbf{Demonstrate the predicted scaling, not merely a constant offset.} A constant offset can be reproduced by any of the mundane loss channels; a measured efficiency whose \emph{ceiling} scales as \(1 - T/T_{\rm res}\) cannot. The discriminating observable is the envelope of \(\eta\) versus \(T/T_{\rm res}\), not a single-point value.
\item
  \textbf{Calibrate the mundane loss channels independently} through complementary measurements that suppress or null each channel separately.
\item
  \textbf{Maintain thermal-like initial conditions for the phonon mode.} As established in Section 3.5, Assumption (A) fails for coherently prepared non-thermal states. The experiment must work with thermally populated phonon modes, not coherently prepared atomic superpositions or squeezed states.
\end{enumerate}

\section{The Astrophysical Regime}\label{the-astrophysical-regime}

The bound (8) applies, in principle, to any process driving displacement against an entropic gradient under Assumption (A). For astrophysical and engineering systems, the regime \(T/T_{\rm res} \ll 1\) generically holds: the local Unruh temperature for any non-extreme gravitational setting is far below the temperature of any plausible energy reservoir.

For a Saturn V launch, the local Unruh temperature corresponding to surface gravity is approximately \(4 \times 10^{-20}\) K, while the combustion chamber temperature is approximately \(3500\) K, giving \(T/T_{\rm res} \sim 10^{-23}\). The bound's prediction \(\eta \leq 1 - 10^{-23}\) is indistinguishable from \(\eta \leq 1\), the trivial second-law statement. The same is true for hot Jupiter atmospheric escape, stellar winds, and active galactic nuclei jets to within factors of order \(10^{10}\) in either direction.

The bound, in this regime, is loose: any system that satisfies the second law will satisfy the bound, regardless of whether the bound's specific functional form captures real physics. We therefore do not present a numerical regression of efficiency margins against \(T/T_{\rm res}\) across astrophysical systems as evidence for the framework. The astrophysical comparison is qualitative context: the bound is compatible with known gravitational phenomenology, but compatibility in a regime where the bound is automatically satisfied is not evidence for the framework's specific functional form.

The decisive test is the BEC experiment of Section 4, in the regime where the bound becomes restrictive.

\section{Falsification Protocol}\label{falsification-protocol}

Because the framework predicts an inequality, only a measured efficiency that \emph{exceeds} the ceiling can falsify the bound outright; a genuine test of the framework's \emph{specific content} rests on the additional attainability claim that a reversible extraction protocol approaches the Carnot ceiling. We therefore distinguish a hard falsification (criterion 1) from consistency checks on the attainability picture (criteria 2--3). The framework is challenged if a controlled BEC analog gravity experiment with \(T/T_{\rm res}\) varied across \([0.05, 0.5]\) produces phonon extraction efficiencies that:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Exceed the bound (8) by more than experimental uncertainty.} Direct violation of the inequality --- the only strict falsification of the stated result.
\item
  \textbf{Show a ceiling with no dependence on \(T/T_{\rm res}\)} as \(g_{\rm eff}\) or \(T_{\rm condensate}\) is varied in a low-loss protocol. A consistency check, not a strict falsification: it would indicate that the holographic-screen substitution does not apply to acoustic analog systems, undermining the transposition assumed in Section 4.1.
\item
  \textbf{Exhibit a low-loss envelope inconsistent with \(\eta = 1 - T/T_{\rm res}\) at the level of experimental error bars.} Also a consistency check on the attainability picture: it would indicate that the analog Unruh identification or the acoustic transposition of the screen-entropy formula does not hold. Because the corrected ceiling is the textbook Carnot form, criterion 3 tests the \emph{physical identifications} \(T \to T_{\rm analog}\), \(T_{\rm res} \to T_{\rm condensate}\), not a novel functional form.
\end{enumerate}

A confirmation outcome --- a low-loss extraction efficiency approaching a ceiling suppressed by approximately 20\% at \(T/T_{\rm res} = 0.2\), with the ceiling tracking \(1 - T/T_{\rm res}\) across the accessible range and surviving independent calibration of mundane loss channels --- would constitute the first laboratory-scale evidence that holographic-screen entropy has thermodynamic content beyond the equilibrium force law.

\section{Limitations}\label{limitations}

We name limitations explicitly. Naming them is part of how this paper aims to be defensible.

\textbf{(a) Assumption (A) is checked numerically only as a near-thermal identity, not derived analytically.} The Lindblad-equation tests of Section 3 reproduce the substitution \(\delta Q = T\,dS\) within numerical error for thermal initial states at \(T/T_{\rm res} \geq 0.20\), but for such states the equality is an identity of the parametrization (Section 3.6), so these runs verify the integrator rather than independently confirming the physical reversibility Assumption (A) posits. They do not constitute an analytical proof. A derivation from first principles --- whether through fluctuation theorems, reaction-coordinate methods, or direct quantum Boltzmann analysis --- would be a substantially stronger result. We do not have one.

\textbf{(b) Assumption (A) fails for non-thermal initial states with macroscopic coherences.} This is a clean negative result from Section 3.5 and we do not minimize it. The framework's experimental validity depends on the BEC experiment being conducted with thermally populated phonon modes, not coherently prepared states.

\textbf{(c) The transposition of Verlinde's screen entropy from real gravitational systems to BEC acoustic analog systems is assumed, not derived.} Analog gravity reproduces the kinematics of curved spacetime in acoustic systems; whether it reproduces holographic-screen thermodynamics is itself the question the proposed experiment addresses. Failure of the experiment may indicate the transposition fails rather than that the underlying gravitational construction fails.

\textbf{(d) The framework predicts an inequality, and its empirical content depends on an attainability claim.} The Carnot bound is saturated only by a reversible (quasi-static, loss-free) extraction protocol, so observing near-saturation requires both the construction to be correct \emph{and} the experiment to approach the reversible limit. Criteria 2--3 of Section 6 are consequently consistency checks on this attainability picture, not strict falsifications --- only a measured efficiency above the ceiling (criterion 1) falsifies the bound itself. A non-equilibrium analog of Jacobson's (1995) derivation --- relating screen entropy production to a geometric or dynamical quantity \emph{as an equation of state} --- would be substantially stronger; we do not have one. The relative-entropy framework developed in the AdS/CFT context (Faulkner et al.~2014) is the most credible existing path toward such a derivation, and we identify it as the natural direction for follow-up theoretical work.

\textbf{(e) At very low temperature ratios (\(T/T_{\rm res} \leq 0.10\)), convergence of the numerical tests is sub-linear in \(dt\).} The substitution still appears to head toward exact equality, but the approach is genuinely slow. Either smaller timesteps or analytic short-time expansions are needed for rigorous validation in this regime. The BEC prediction at \(T/T_{\rm res} = 0.20\) sits on the safe side of this boundary, but extensions of the experiment to \(T/T_{\rm res} < 0.10\) would require further numerical or analytical work.

\textbf{(f) The astrophysical comparison in Section 5 is qualitative.} It demonstrates compatibility, not support. The regime \(T/T_{\rm res} \ll 1\) makes the bound loose and the comparison non-discriminating.

\textbf{(g) The experimental sensitivity required to resolve the predicted 20\% suppression has not been formally established.} Section 4.3 indicates that the operational definition of \(\eta\) corresponds to a measured quantity in published BEC experiments. Whether the suppression is resolvable above the systematic error floor of any specific experimental setup requires direct dialogue with experimental groups working in this regime.

\textbf{(h) Earlier drafts of this work invoked entanglement-erasure interpretations and quantum Landauer-principle language. We have not retained these here} because the language was not doing technical work in the derivation. If equation (8) is empirically confirmed, an entanglement-theoretic re-derivation may eventually be possible, but the present construction does not require it.

\section{Conclusions}\label{conclusions}

We have applied standard non-equilibrium thermodynamics to a Verlinde-style holographic screen entropy and obtained, under one explicit thermodynamic assumption (Assumption A of Section 2.3), an efficiency bound:

\[\eta \leq 1 - \frac{T}{T_{\rm res}}.\]

The substitution that defines Assumption (A) has been examined numerically across five settings (Section 3) using QuTiP Lindblad simulations. For thermal initial states it is an algebraic identity that the simulations reproduce within numerical error in the dt → 0 limit --- a consistency check on the integrator rather than an independent test --- and it fails for non-thermal initial states with macroscopic coherences, a clean limit on the framework's scope. An independent test of Assumption (A), in a model where the substitution is not guaranteed by the state's structure, remains open.

We have not modified Newtonian or general-relativistic gravity. We have not derived a new equation of motion. We have not unified entropy and gravity in the sense of Ryu--Takayanagi or the broader entanglement-as-geometry program (Van Raamsdonk 2010; Faulkner et al.~2014). We claim only the following: that the bound, within its stated scope, predicts a 20\% efficiency suppression in BEC analog gravity at \(T/T_{\rm res} = 0.2\), distinguishable from naive energetic accounting. This is a falsifiable laboratory prediction in a regime where current experimental capabilities can plausibly access it.

The framework's empirical content stands or falls on this prediction. The astrophysical compatibility of Section 5 is illustrative context, not evidence. A successful BEC test would establish that holographic-screen entropy of the Verlinde (2011) type has thermodynamic content beyond the equilibrium force law --- a small but non-trivial empirical result. A failed test would falsify the construction or its acoustic transposition. Either outcome is informative.

Future theoretical work should address: (i) an analytical derivation of Assumption (A) generalizing the Lindblad-model evidence of Section 3; (ii) the validity of transposing screen entropy to acoustic analog systems; (iii) the relation between this non-equilibrium inequality and a putative non-equilibrium equation of state generalizing Jacobson (1995). Future experimental work should establish, in dialogue with active BEC analog gravity groups, the specific implementation and signal-to-noise budget required to distinguish the predicted 20\% suppression from mundane experimental losses.

\section*{Data Availability}\label{data-availability}

As of this version, the simulation scripts, raw output logs, and convergence diagnostics for the Section 3 runs are \textbf{not yet publicly archived} --- the Zenodo deposit currently holds the paper only, and the companion Windstorm Labs repository is a placeholder pending upload. The numerical method for each test in Section 3 is specified in full in the text, and the scripts will be released in the Labs repository (github.com/Windstorm-Labs/phonon-extraction-bound) and mirrored to a new Zenodo version. Until then the paper is the authoritative description of the numerics.

The simulations were run using QuTiP 5.2.3 in standard Python 3.12 environments; total compute cost across all reported tests is approximately 30 minutes on a standard CPU.

\section*{Methodology Note}\label{methodology-note}

This paper was developed iteratively with assistance from large language models (Claude, Grok, Gemini, ChatGPT, Perplexity) used as adversarial review and computational assistance tools. The numerical experiments in Section 3 were executed by code-execution-enabled instances of these models running real Python scripts in sandboxed environments; no quantitative result reported here was generated without code execution, and at least two independent runs were conducted to cross-check key findings; the scripts themselves are not yet archived (see Data Availability), so this process is described rather than currently reproducible by the reader. Earlier drafts of this work contained claims (numerical regressions over astrophysical datasets, modified-gravity force-law extensions, cosmological connections) that were progressively removed during the review process when they could not be defended. The methodology is documented for transparency and to invite scrutiny of the inputs that produced this manuscript.

\subsection*{Data, Code \& Resources}
\noindent Manuscript source and archived record: \url{https://github.com/Windstorm-Institute/phonon-extraction-bound} (Zenodo: \href{https://doi.org/10.5281/zenodo.20014390}{10.5281/zenodo.20014390}). Experiments, datasets, and analysis code: \url{https://github.com/Windstorm-Labs/phonon-extraction-bound}. Windstorm Institute: \url{https://windstorminstitute.org}. Windstorm Labs: \url{https://windstormlabs.com}.

\section*{References}\label{references}

Barceló, C., Liberati, S. \& Visser, M. (2005). Analogue gravity. \emph{Living Reviews in Relativity}, \textbf{8}, 12.

Bekenstein, J. D. (1973). Black holes and entropy. \emph{Physical Review D}, \textbf{7}, 2333.

Faulkner, T., Guica, M., Hartman, T., Myers, R. C. \& Van Raamsdonk, M. (2014). Gravitation from entanglement in holographic CFTs. \emph{Journal of High Energy Physics}, \textbf{2014}(3), 51.

Hawking, S. W. (1975). Particle creation by black holes. \emph{Communications in Mathematical Physics}, \textbf{43}, 199.

Jacobson, T. (1995). Thermodynamics of spacetime: the Einstein equation of state. \emph{Physical Review Letters}, \textbf{75}, 1260.

Johansson, J. R., Nation, P. D. \& Nori, F. (2013). QuTiP 2: A Python framework for the dynamics of open quantum systems. \emph{Computer Physics Communications}, \textbf{184}, 1234.

Lahav, O., Itah, A., Blumkin, A., Gordon, C. \& Steinhauer, J. (2010). Realization of a sonic black hole analog in a Bose--Einstein condensate. \emph{Physical Review Letters}, \textbf{105}, 240401.

Landauer, R. (1961). Irreversibility and heat generation in the computing process. \emph{IBM Journal of Research and Development}, \textbf{5}, 183.

Muñoz de Nova, J. R., Golubkov, K., Kolobov, V. I. \& Steinhauer, J. (2019). Observation of thermal Hawking radiation and its temperature in an analogue black hole. \emph{Nature}, \textbf{569}, 688.

Ryu, S. \& Takayanagi, T. (2006). Holographic derivation of entanglement entropy from AdS/CFT. \emph{Physical Review Letters}, \textbf{96}, 181602.

Steinhauer, J. (2016). Observation of quantum Hawking radiation and its entanglement in an analogue black hole. \emph{Nature Physics}, \textbf{12}, 959.

Unruh, W. G. (1976). Notes on black-hole evaporation. \emph{Physical Review D}, \textbf{14}, 870.

Unruh, W. G. (1981). Experimental black-hole evaporation? \emph{Physical Review Letters}, \textbf{46}, 1351.

Van Raamsdonk, M. (2010). Building up spacetime with quantum entanglement. \emph{General Relativity and Gravitation}, \textbf{42}, 2323.

Verlinde, E. (2011). On the origin of gravity and the laws of Newton. \emph{Journal of High Energy Physics}, \textbf{2011}(4), 29.

\end{document}
